Comparing Generative Pre-trained Transformer Models to Physicians for Assigning Causes of Death from Verbal Autopsy: A Case Study of 6939 Deaths in Sierra Leone from 2019-2022

### 2024-07-07 manuscript/manuscript.tex
Finish conclusion

This study evaluates the performance of GPT-3.5, GPT-4, InterVA-5, and InSilicoVA models compared to physicians for assigning CODs for 6939 VA records in Sierra Leone (2019-2022). At the population level, all models had CSMF accuracies of 0.7-0.79. At the individual level, GPT-4 had the best performance (0.61 PCCC), followed by GPT-3.5 (0.58 PCCC), InSilicoVA (0.44 PCCC), and InterVA-5 (0.45 PCCC). When evaluating performance by COD, 7 of 19, 4 of 10, and 1 of 7 adult (12-69 years), child (28 days to 11 years),
and neonatal (under 28 days) CODs respectively had models with $\geq$0.8 PCCC (see Table \ref{tab-perf-geq80}). Performance decreased from 0.72 to 0.59 PCCC as adults aged, and increased from 0.5 to 0.7 PCCC as children and neonates developed. Thus, GPT and InSilicoVA models were comparable to physicians for particular CODs (e.g. maternal conditions, injuries, diarrhoeal diseases, pneumonia), but not across age ranges. Advantages of GPT (and InSilicoVA for some CODs) models include being highly scalable and available, which allows the suggestion of multiple alternative COD assignments, and reduction of time spent on non-complex cases to assist physicians in COD assignment. In addition, GPT models provide flexible natural language input and output, capturing latent patterns (e.g. health-seeking and social issues) that potentially lead to their overall high performance compared to InterVA-5 and InSilicoVA. However, GPT models do not always assign the same COD for the same record on multiple runs, are trained on data from a specific time period, and require large computing infrastructure, leading to disadvantages in reliability of COD assignments, timeliness of up-to-date information, and data privacy issues. Limitations of this study were small sample sizes in relation to rarer CODs and other national level studies, need for more comprehensive exploration of disagreed and misclassified records, and lack of experimentation of model parameters, output consistency, and multiple COD assignments. Research opportunities include refining GPT models using prompt engineering and custom models, and future work towards computer assisted verbal autopsy, where GPT and other models are used to assist physician COD assignment by offering multiple alternative assignments, improving physician agreement on COD assignment and reducing ill-defined deaths.

>>>>

This study evaluates the performance of GPT-3.5, GPT-4, InterVA-5, and InSilicoVA models compared to physicians for assigning CODs for 6939 VA records in Sierra Leone (2019-2022). At the population level, all models had CSMF accuracies of 0.7-0.79. At the individual level, GPT-4 had the best performance (0.61 PCCC), followed by GPT-3.5 (0.58 PCCC), InSilicoVA (0.44 PCCC), and InterVA-5 (0.45 PCCC). When evaluating performance by COD, 7 of 19, 4 of 10, and 1 of 7 adult (12-69 years), child (28 days to 11 years),
and neonatal (under 28 days) CODs respectively had models with $\geq$0.8 PCCC (see Table \ref{tab-perf-geq80}). Performance decreased from 0.72 to 0.59 PCCC as adults aged, and increased from 0.5 to 0.7 PCCC as children and neonates developed. Thus, GPT and InSilicoVA models were comparable to physicians for particular CODs (e.g. maternal conditions, injuries, diarrhoeal diseases, pneumonia), but not across age ranges. Advantages of GPT (and InSilicoVA for some CODs) models include being highly scalable and available, which allows the suggestion of multiple alternative COD assignments, and reduction of time spent on non-complex cases to assist physicians in COD assignment. In addition, GPT models provide flexible natural language input and output, capturing latent patterns (e.g. health-seeking and social issues) that potentially lead to their overall high performance compared to InterVA-5 and InSilicoVA. However, GPT models do not always assign the same COD for the same record on multiple runs, are trained on data from a specific time period, and require large computing infrastructure, leading to disadvantages in reliability of COD assignments, timeliness of up-to-date information, and data privacy issues. Limitations of this study were small sample sizes in relation to rarer CODs and other national level studies, need for more comprehensive exploration of disagreed and misclassified records, and lack of experimentation of model parameters, output consistency, and multiple COD assignments. Research opportunities include refining GPT models using prompt engineering and custom models, and future work towards computer assisted verbal autopsy, where GPT and other models are used to assist physician COD assignment by offering multiple alternative assignments, improving physician agreement on COD assignment and reducing ill-defined deaths. GPT-4, InterVA-5, and InSilicoVA has been integrated into future rounds of the HEAl-SL study to assist physicians with alternative COD assignments. Future work in evaluating the effectiveness of these models to reduce disagreements among physicians and ill-defined deaths will be a step forward in more accurate and efficient VA systems.
